---
title: "Hypothesis testing strategy for DLNMs"
output: html_notebook
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, paged.print = FALSE)
options(scipen = 4)
```

*Last compiled: `r Sys.Date()`*

```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(here)
library(conflicted)
library(mgcv)
library(dlnm)
library(performance)
library(bbmle)


conflict_prefer("filter", "dplyr")
conflict_prefer("here", "here")
conflict_prefer("lag", "dplyr")
conflict_prefer("select", "dplyr")

source(here("R", "crossbasis.r"))
```

```{r include=FALSE}
hoppers <- read_rds(here("data", "cleaned", "hopper-weather.rds"))
```

```{r include=FALSE}
m_hopper_temp_max <- 
  gam(hoppers ~
        s(Q_temp_max, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
        ) +
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers2,
      gamma = 1.2,
      method = "REML")
```

# Question

I've thought about a way to do hypothesis testing for the manuscript that appeases both reviewers and me, but I'm not sure how it is best to include. It's 3 tests x 4 variables (precip, min temp, mean temp, max temp). I'm thinking it might be appropriate to include as a table, but would love to hear your thoughts.

# Background

Functional linear models (FLMs) are linear models (regressions) that include a continuous function as a covariate. For example, rather than including temperature as a covariate, a FLM might include a function of temperature over time as a covariate. Distributed lag non-linear models (DLNMs) are FLMs that model non-linear and potentially lagged effects as two-dimensional smooth functions. This allows for modeling of non-linear effects of the variable of interest as well as a non-linear effect through time. This 2-dimensional smooth is referred to as a "crossbasis" by Gasparrini et al.

I'm implementing DLNMs using a generalized additive model (GAM) framework. Basically the GAM figures out the appropriate smoothing penalties for the crossbasis. One problem with this is that p-values are known to be too low for penalized smooths in GAMs. On a related note, reviewer 2 notes:

> One thing that might help would be to contrast this statistical approach [DLNM] with another, I think more common, approach to inferring the importance of different time lags, which is to consider only one lag at a time and use model selection methods to determine which lags produce the model that minimizes the AIC or explains the most variance.

So as a sanity check on the p-values obtained from DLNMs, and to address the reviewer's concerns, I propose testing three explicit hypotheses with the DLNMs inspired by Swihart et al. [-@swihartRestrictedLikelihoodRatio2014].

# Hypotheses to test

1.  Is *smooth* function necessary?

    -   H~A~: Full model including crossbasis smooth for weather variable over 15 days lag.
    -   H~0~: replace crossbasis term with mean of temperature in past 15 days

2.  Is there an effect of temperature?

    -   H~A~: Same as H~A~ from \#1 above
    -   H~0~: Remove the weather variable term entirely

3.  Is there a single lagged effect?

    -   H~A~: Use the weather variable at a lag with maximum effect suggested by the DLNM (e.g. mean temperature 10 days ago) as a parameteric (not smoothed) term.
    -   H~0~: Same as H~0~ from \#2 above

I'll now go through an example with the effects of mean temperature on leafhopper densities. Terms in the full model include:

-   `s(Q_temp_mean, L, bs = "cb")` - This is the crossbasis smooth of mean daily temperature over the past 15 days

-   `s(day_post, bs = "cr")` - A potentially non-linear effect of days post harvest

-   `s(counter, bs = "re")` - A random effect of who did the data collection (three observers)

-   `s(interharvest_id, bs = "re")` - A random effect of harvest (date range x field ID).

-   `s(plant, bs = "re")` - A random effect of plant ID (to account for repeated measures).

## Is the crossbasis smooth necessary?

```{r echo=FALSE}
#calculate mean of past 15 days
hoppers2 <-
  hoppers %>%
  rowwise() %>% 
  mutate(mean_temp_mean = mean(Q_temp_mean))

m <- 
  gam(hoppers ~
        s(Q_temp_mean, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
        ) +
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers2,
      gamma = 1.2,
      method = "REML")
m1 <- 
  gam(hoppers ~
        mean_temp_mean +
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers2,
      gamma = 1.2,
      method = "REML")
test1 <- anova.gam(m, m1, test = "Chisq")
test1
```

Including a crossbasis smooth significantly improves the model over simply averaging temperatures over the previous 15 days from an observation (p = `r test1[["Pr(>Chi)"]][2]`)

## Is there an effect of weather term?

```{r}
m2 <- 
  gam(hoppers ~
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = `(offset)`, #this is how the offset shows up in model.frame(m)
      data = model.frame(m), #to ensure its the same data
      gamma = 1.2,
      method = "REML")
test2 <- anova.gam(m, m2, test = "Chisq") 
test2
```

Mean daily temperature over the previous 15 days has a significant effect on leafhopper abundance (p = `r test2[["Pr(>Chi)"]][2]`).

## Effect of specific lag

According to the crossbasis plot (not shown here), the mean daily temperature 10 days ago had the strongest effect on leafhopper density.

```{r}
m3 <- 
  gam(hoppers ~
        Q_temp_mean[, 10] +
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = `(offset)`, #this is how the offset shows up in model.frame(m)
      data = model.frame(m), #to ensure its the same data
      gamma = 1.2,
      method = "REML")

test3 <- anova.gam(m3, m2, test = "Chisq")
test3
```

The mean daily temperature 10 days ago alone did not have a significant effect on leafhopper density (p = `r test3[["Pr(>Chi)"]][2]`). Because the other two tests are significant, this supports a cumulative effect of temperature (i.e. the "distributed" in DLNM) rather than an effect of a specific point in time prior to the observed leafhopper density.

# References
