---
title: "EDA and descriptive figures"
author: "Eric R. Scott"
date: "2022-03-26"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

#TODO:

- ~~Figure 1 should include entire time range used for data~~
- ~~Add grey rectangle for missing values in weather?~~
- ~~Calculate mean daily min, mean, max temperatures for entire range.~~
- ~~Re-do precip calculations (mean, % of rainy days) for entire range.~~
- Concurvity models broke.  Just need to update variable names, I think.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Last compiled: `r Sys.Date()`*

```{r packages, include=FALSE}
library(tidyverse)
library(lubridate)
library(here)
library(patchwork)
library(scales)
library(mgcv)
library(conflicted)
library(tsibble)
conflict_prefer("filter", "dplyr")
```

# Purpose

Make a plot that shows leafhopper densities over time with harvests and maybe weather.

Also a supplementary plot showing concurvity of temp and precip.

# Load Data

```{r data, echo=TRUE}
df <- read_rds(here("data", "cleaned", "hopper-weather.rds"))
weather_full <- read_rds(here("data", "cleaned", "weather_lags.rds"))
```

# Filter to include only dates used
Right now, I'm using weather up to 15 days prior to observations

```{r}
#first observsation:
min(df$date)
#first date for weather:
min(df$date) - days(15)

weather_full <-
  weather_full %>% 
  filter(date >= min(df$date) - days(15)) %>% 
  #don't need matrix-colums
  select(-starts_with("Q_"), -starts_with("L")) %>% 
  as_tsibble(index = date) %>% 
  fill_gaps()
```

# Summary tables

```{r}
df %>% 
  group_by(interharvest_id) %>% 
  summarize(length = last(date) - first(date),
            mean_temp = mean(temp_mean),
            mean_precip = mean(precip, na.rm = TRUE),
            num_rainy = sum(precip > 0, na.rm = TRUE)/n(),
            median_precip = median(precip, na.rm = TRUE),
            mean_density = mean(hoppers/n_leaves))

df %>%
    summarize(length = last(date) - first(date),
            mean_temp_mean = mean(temp_mean),
            mean_precip = mean(precip, na.rm = TRUE),
            max_precip = max(precip, na.rm = TRUE),
            mean_density = mean(hoppers/n_leaves))
  
```

# Leafhopper Density Plot

## Summarize data for plotting

```{r}
plotdf <- df %>% 
  group_by(date, field) %>% 
  summarize(density = mean(hoppers / n_leaves),
            harvest = first(interharvest_id)) %>% 
  filter(harvest != "A4")
```


## Plot

add lines/colors for harvests?
add panel for temp
add panel for precip

```{r}
densityplot <-
  ggplot(plotdf, aes(x = date, y = density, group = field, color = harvest)) + 
  geom_line(size = 1, lineend = "round") +
  theme_bw() +
  labs(x = "Date", y = "Leafhopper Density (insects/leaf)", color = "Harvest") +
  scale_x_date(date_breaks = "3 days", labels = label_date(format = "%b %d"), limits = range(weather_full$date)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
densityplot
```


```{r}
ggplot(df, aes(x = date, y = hoppers/n_leaves, group = plant)) +
  geom_line(alpha = 0.1) +
  geom_line(data = plotdf, aes(y = density, group = field, color = harvest), size = 1) +
  theme_bw() +
  labs(x = "Date", y = "Leafhopper Density (insects/leaf)", color = "Harvest") +
  scale_x_date(date_breaks = "1 week", labels = label_date(format = "%b %d"))
# ggsave("test.png")
```

# Figure 1

```{r}
max(weather_full$precip, na.rm = TRUE)
precip_plot <-
  ggplot(weather_full, aes(x = date, y = precip)) +
  geom_col(fill = "darkblue") +
  theme_bw() +
  scale_y_continuous("Precipitation (mm)") +
  scale_x_date(date_breaks = "3 days", labels = label_date(format = "%b %d")) +
  theme(axis.text.x.bottom = element_blank(),
        axis.title.x.bottom = element_blank())
```


```{r}
which(is.na(weather_full$temp_mean))
min(weather_full$temp_min, na.rm = TRUE)
max(weather_full$temp_max, na.rm = TRUE)
temp_plot <-
  ggplot(weather_full, aes(x = date)) +
  geom_line(aes(y = temp_mean), color = "red") +
  geom_line(aes(y = temp_min), color = "red", lty = "dashed") +
  geom_line(aes(y = temp_max), color = "red", lty = "dashed") +
  theme_bw() +
  scale_y_continuous("Temperature (ºC)") +
  scale_x_date(date_breaks = "3 days", labels = label_date(format = "%b %d")) +
  theme(axis.text.x.bottom = element_blank(),
        axis.title.x.bottom = element_blank())
```

```{r}
plot1 <- precip_plot/temp_plot/densityplot +
  plot_layout(heights = c(2,2,3.5)) +
  plot_annotation(tag_levels = "A") &
  scale_x_date(date_breaks = "3 days",
               labels = label_date(format = "%b %d"),
               limits = range(weather_full$date),
               expand = c(0,0)) &
  theme(text = element_text(size = 9))
plot1
ggsave(here("figs", "figure 1.png"), plot1, width = 165, height = 150, units = "mm")
ggsave(here("figs", "figure 1.eps"), plot1, width = 165, height = 150, units = "mm")
```
Grey bars represent two days of missing weather data.

# Correlation / concurvity supplementary plot

```{r}
shootdf <- read_rds(here("data", "cleaned", "shoots.rds"))
library(dlnm)
```

```{r}
m_shoots <- 
  gam(growth ~ 
        #cross-basis function for temperature
        s(Q_temp_mean, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
          ) +
        #cross-basis function for precipitation
        s(Q_precip, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
          ) +
        s(day_post, bs = "cr") + s(diameter, bs = "cr") + #co-variates
        s(interharvest_id, bs = "re") + s(plant, bs = "re"), #random effects
      data = shootdf,
      gamma = 1.2,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      method = "REML")

summary(m_shoots)
```

```{r eval = FALSE}
#concurvity() doesn't work with NAs in the data.
#might also not work because a term has < 1 edf?
concurvity(m_shoots, full = TRUE)
concurvity(m_shoots, full = FALSE)
```

Precip and temp are concurve meaning they might be redundant in the model.  This makes sense as temperature and precip are generally correlated to some extent.

```{r}
concurvity_p <- 
  ggplot(weather_full, aes(y = temp_mean, x = precip)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "gam", level = 0.95) +
  labs(x = "Precipitation (mm)", y = "Mean Temperature (ºC)") +
  theme_bw()
concurvity_p
ggsave(here("figs", "plotS1.png"), concurvity_p)
ggsave(here("figs", "plotS1.tiff"), concurvity_p, width = 165, units = "mm")
```

```{r}
test <-gam(temp_mean ~ s(precip, bs = "cs", k = 14), data = weather_full)
gam.check(test)
anova.gam(test)
plot(test)
```


# Descriptives

```{r}
weather_full %>% 
  as_tibble() %>% 
  summarize(across(-date, list(mean = ~mean(.x, na.rm = TRUE),
                               sd = ~sd(.x, na.rm = TRUE))),
            rainy_days_percent = sum(precip >0, na.rm = TRUE)/n())
```

```{r}
df %>% 
  summarize(hopper_density = mean(hoppers/n_leaves),
            hopper_density_sd = sd(hoppers/n_leaves))
```


