---
title: "DLNM plots"
author: "Eric R. Scott"
date: "2020-12-14"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(mgcv)
library(dlnm)
library(patchwork)
library(visibly)

source(here("R", "crossbasis.r"))

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

What is the goal of this notebook?

# Load Data

```{r data, echo=TRUE}
hoppers <- read_rds(here("data", "cleaned", "hopper-weather.rds"))
shootdf <- read_rds(here("data", "cleaned", "shoots.rds"))

m_hopper_precip <- read_rds(here("analysis", "models", "m_hopper_precip.rds"))
m_hopper_temp_max  <- read_rds(here("analysis", "models", "m_hopper_temp_max.rds"))
m_hopper_temp_mean <- read_rds(here("analysis", "models", "m_hopper_temp_mean.rds"))
m_hopper_temp_min  <- read_rds(here("analysis", "models", "m_hopper_temp_min.rds"))
```

# Leafhopper Plots

## Covaraiate

Days post-harvest

```{r}
d <- list("Mean Temperature" = m_hopper_temp_mean,
         "Minimum Temperature" = m_hopper_temp_min,
         "Maximum Temperature" = m_hopper_temp_max,
         "Precipitation" = m_hopper_precip) %>%
  map(~plot_gam(.x, main_var = day_post))

dph_plot <-
  d %>%
  map_df(~pluck(.x,"data"), .id = "model") %>% 
  ggplot(aes(x = value, y = fit, color = model)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), 
              fill = "grey", linetype = 0, alpha = 0.5) +
  geom_line(size = 1) +
  geom_rug(data = hoppers, aes(day_post), inherit.aes = FALSE, alpha = 0.03) +
  labs(x = "Days post harvest", y = "Leafhoppers / leaf", color = "Model") +
  theme_bw() +
  theme(aspect.ratio = 1)
dph_plot
ggsave(here("figs", "hopper_covariates.png"), dph_plot, width = 165, height = 130, units = "mm")
```

## Crossbasis plots

### Generate plot data

The `cb_margeff()` function sets covariates to their mean (or reference level if they are factors) and then runs `predict()` on a range of temperature histories (e.g. `Q_temp_mean, L`).

```{r}
ht_pd <- cb_margeff(m_hopper_temp_mean, Q_temp_mean, L) 
htmin_pd <- cb_margeff(m_hopper_temp_min, Q_temp_min, L)
htmax_pd <- cb_margeff(m_hopper_temp_max, Q_temp_max, L)
hp_pd <- cb_margeff(m_hopper_precip, Q_precip, L) 
```


Create a common scale for fitted values so colors are comparable.

```{r}
hopper_lim <- c(min(exp(ht_pd$fitted), 
                    exp(htmin_pd$fitted),
                    exp(htmax_pd$fitted), 
                    exp(hp_pd$fitted)),
                max(exp(ht_pd$fitted),
                    exp(htmin_pd$fitted),
                    exp(htmax_pd$fitted),
                    exp(hp_pd$fitted)))

contour_binwidth <- 0.003 #distance between contour lines (in leafhoppers/leaf)
```

### Mean Temperature cross-basis

```{r}
hopper_temp_raster <-
  ggplot(ht_pd,
         aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  geom_raster(data = filter(ht_pd, min_dist >= 0.1), fill = "white", alpha = 0.4) +
  geom_contour(color = "grey50", size = 0.4, binwidth = contour_binwidth) +
  scale_fill_viridis_c("Leafhoppers\nper leaf",
                       option = "A",
                       limits = hopper_lim) + 
  scale_y_continuous("Lag (days)", expand = c(0,0)) + 
  scale_x_continuous("Daily Mean Temperature (ºC)", expand = c(0,0))
hopper_temp_raster
```

WHere is steepest slope?

```{r}
ht_pd %>% 
  filter(fitted == max(fitted))
#9 days lag
```


### Minimum Temperature cross-basis

```{r}
hopper_min_temp_raster <-
  ggplot(htmin_pd,
         aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  geom_raster(data = filter(htmin_pd, min_dist >= 0.1), fill = "white", alpha = 0.4) +
  geom_contour(color = "grey50", size = 0.4, binwidth = contour_binwidth) +
  scale_fill_viridis_c("Leafhoppers\nper leaf",
                       option = "A", 
                       limits = hopper_lim) + 
  scale_y_continuous("Lag (days)", expand = c(0,0)) + 
  scale_x_continuous("Daily Minimum Temperature (ºC)", expand = c(0,0))
hopper_min_temp_raster
```
Where is steepest slope?

```{r}
htmin_pd %>% 
  filter(min_dist < 0.1) %>% 
  filter(x > 12) %>% 
  # filter(x < 22) %>% 
  filter(fitted == max(fitted))
#1 days lag and maybe also 10 days?
```

### Maximum temperature cross-basis

```{r}
hopper_max_temp_raster <-
  ggplot(htmax_pd,
         aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  geom_raster(data = filter(htmax_pd, min_dist >= 0.1), fill = "white", alpha = 0.4) +
  geom_contour(color = "grey50", size = 0.4, binwidth = contour_binwidth) +
  scale_fill_viridis_c("Leafhoppers\nper leaf",
                       option = "A",
                       limits = hopper_lim) + 
  scale_y_continuous("Lag (days)", expand = c(0,0)) + 
  scale_x_continuous("Daily Maximum Temperature (ºC)", expand = c(0,0))
hopper_max_temp_raster
```
Where is steepest slope?

```{r}
htmax_pd %>% 
  filter(min_dist < 0.1) %>% 
  filter(fitted == max(fitted))
#10 days lag
```

### Precipitation cross-basis

```{r}
hopper_precip_raster <-
  ggplot(hp_pd,
         aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  geom_raster(data = filter(hp_pd, min_dist >= 0.1), fill = "white", alpha = 0.4) +
  geom_contour(color = "grey50", size = 0.4, binwidth = contour_binwidth) +
  scale_fill_viridis_c("Leafhoppers\nper leaf",
                       option = "A",
                       limits = hopper_lim) + 
  scale_y_continuous("Lag (days)", expand = c(0,0)) + 
  scale_x_continuous("Daily Precipitation (mm)", expand = c(0,0))
hopper_precip_raster
```

Where is steepest slope?

```{r}
hp_pd %>% 
  filter(min_dist < 0.1) %>% 
  filter(fitted == max(fitted))
#12 days lag
```

### Combine raster plots

```{r}
hopper_cb <-
  hopper_min_temp_raster +
  hopper_temp_raster +
  hopper_max_temp_raster +
  hopper_precip_raster +
  plot_layout(guides = "collect") &
  coord_cartesian(ylim = c(1,15)) &
  plot_annotation(tag_levels = "A") &
  theme_classic() &
  theme(text = element_text(size = 10))
hopper_cb
ggsave(here("figs", "hopper_crossbasis.png"), hopper_cb, width = 165, height = 120, units = "mm")
ggsave(here("figs", "figure 2.eps"), hopper_cb, width = 165, height = 120, units = "mm")
```

Try adding a rug for weather data

```{r}
weather <- read_rds(here("data", "cleaned", "weather_lags.rds"))
weather <- weather %>% 
  filter(date > min(hoppers$date) - lubridate::days(15),
         date != max(hoppers$date)) #last day wouldn't show up in lags for any observations
range(weather$date)
range(hoppers$date)

hopper_cb_rug <-
  hopper_min_temp_raster +
  geom_rug(data = weather, inherit.aes = FALSE, aes(x = temp_min), color = "white") +
hopper_temp_raster +
  geom_rug(data = weather, inherit.aes = FALSE, aes(x = temp_mean), color = "white") +
hopper_max_temp_raster +
  geom_rug(data = weather, inherit.aes = FALSE, aes(x = temp_max), color = "white") +
hopper_precip_raster +
  geom_rug(data = weather, inherit.aes = FALSE, aes(x = precip), color = "white")

hopper_cb_rug <- 
  hopper_cb_rug +
  plot_layout(guides = "collect") &
  coord_cartesian(ylim = c(1,15)) &
  plot_annotation(tag_levels = "A") &
  theme_classic() &
  theme(text = element_text(size = 10))

ggsave(here("figs", "hopper_crossbasis_rug.png"), hopper_cb_rug, width = 165, height = 120, units = "mm")
```


# Shoot Growth Plots


```{r}
m_shoot_precip <- read_rds(here("analysis", "models",    "m_shoot_precip.rds"))
m_shoot_temp_max <- read_rds(here("analysis", "models",  "m_shoot_temp_max.rds"))
m_shoot_temp_mean <- read_rds(here("analysis", "models", "m_shoot_temp_mean.rds"))
m_shoot_temp_min <- read_rds(here("analysis", "models",  "m_shoot_temp_min.rds"))
```

## Covariates:

Extract data from GAMs

```{r}
covar_plot_data <-
  list(
    "Mean Temperature" = m_shoot_temp_mean,
    "Minimum Temperature" = m_shoot_temp_min,
    "Maximum Temperature" = m_shoot_temp_max,
    "Precipitation" = m_shoot_precip
  ) %>%
  map(~plot_gam(.x, main_var = vars(day_post, diameter))) %>% 
  map_df(~pluck(.x,"data"), .id = "model")
```

Days post-harvest

```{r}
dph_plot <-
  covar_plot_data %>% 
  filter(term == "day_post") %>% 
  ggplot(aes(x = value, y = fit, color = model)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), 
              fill = "grey", linetype = 0, alpha = 0.5) +
  geom_line(size = 1) +
  geom_rug(data = shootdf,
           aes(day_post),
           inherit.aes = FALSE,
           alpha = 0.03) +
  scale_y_continuous(limits = c(0, 0.6)) +
  labs(x = "Days post harvest", y = "Shoot growth (cm/day)", color = "Model") +
  theme_bw()
dph_plot
```

```{r}
dia_plot <- 
  covar_plot_data %>% 
  filter(term == "diameter") %>% 
  ggplot(aes(x = value, y = fit, color = model)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), 
              fill = "grey", linetype = 0, alpha = 0.5) +
  geom_line(size = 1) +
  geom_rug(data = shootdf,
           aes(diameter),
           inherit.aes = FALSE,
           alpha = 0.03) +
  scale_y_continuous(limits = c(0, 0.6)) +
  labs(x = "Stem Diameter (mm)", y = "Shoot growth (cm/day)", color = "Model") +
  theme_bw()
dia_plot
```


Shoots with larger diameters grow faster.  Shoot growth rates vary through time in a more complex manner with the fastest growth happening about 10 days post harvest and the slowest growth around 25 days post harvest.


```{r}
covars <-
  dia_plot + 
  (dph_plot +
     theme(axis.title.y = element_blank(),
           axis.text.y = element_blank(),
           axis.ticks.y = element_blank())) + 
  plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r}
ggsave(here("figs", "shoot_covariates.png"), covars, width = 165, height = 80, units = "mm")
ggsave(here("figs", "figure 4.tiff"), covars, width = 165, height = 80, units = "mm")
```

## Crossbasis plots
### Generate data

```{r}
st_pd <- cb_margeff(m_shoot_temp_mean, Q_temp_mean, L) 
stmin_pd <- cb_margeff(m_shoot_temp_min, Q_temp_min, L)
stmax_pd <- cb_margeff(m_shoot_temp_max, Q_temp_max, L)
sp_pd <- cb_margeff(m_shoot_precip, Q_precip, L)
```

Uniform scale across plots:

```{r}
growth_lims <- c(min(st_pd$fitted, 
                     sp_pd$fitted,
                     stmin_pd$fitted, 
                     stmax_pd$fitted),
                 max(st_pd$fitted,
                     sp_pd$fitted,
                     stmin_pd$fitted,
                     stmax_pd$fitted))
contour_binwidth <- 0.025 #distance between contour lines (in cm)
```

### Mean Temperature cross-basis

Raster version:

```{r}
shoot_temp_raster <-
  ggplot(st_pd, aes(x = x, y = lag, z = fitted, fill = fitted)) +
  geom_raster(interpolate = TRUE) +
  geom_raster(data = filter(st_pd, min_dist >= 0.1), fill = "white", alpha = 0.4) +
  geom_contour(color = "grey50", size = 0.4, binwidth = contour_binwidth) +
  scale_fill_viridis_c("Shoot growth\n(cm/day)", option = "A", limits = growth_lims) + 
  scale_y_continuous("Lag (days)", expand = c(0,0)) + 
  scale_x_continuous("Daily Mean Temperature (ºC)", expand = c(0,0)) 
shoot_temp_raster
```


### Min temperature cross-basis

```{r}
shoot_min_temp_raster <-
  ggplot(stmin_pd,
         aes(x = x, y = lag, z = fitted, fill = fitted)) +
  geom_raster(interpolate = TRUE) +
  geom_raster(data = filter(stmin_pd, min_dist >= 0.1), fill = "white", alpha = 0.4) +
  geom_contour(color = "grey50", size = 0.4, binwidth = contour_binwidth) +
  scale_fill_viridis_c("Shoot growth\n(cm/day)", option = "A", limits = growth_lims) + 
  scale_y_continuous("Lag (days)", expand = c(0,0)) + 
  scale_x_continuous("Daily Minimum Temperature (ºC)", expand = c(0,0)) 
shoot_min_temp_raster
```

### Max temperature cross-basis

```{r}
shoot_max_temp_raster <-
  ggplot(stmax_pd,
         aes(x = x, y = lag, z = fitted, fill = fitted)) +
  geom_raster(interpolate = TRUE) +
  geom_raster(data = filter(stmax_pd, min_dist >= 0.1), fill = "white", alpha = 0.4) +
  geom_contour(color = "grey50", size = 0.4, binwidth = contour_binwidth) +
  scale_fill_viridis_c("Shoot growth\n(cm/day)", option = "A", limits = growth_lims) + 
  scale_y_continuous("Lag (days)", expand = c(0,0)) + 
  scale_x_continuous("Daily Maximum Temperature (ºC)", expand = c(0,0))
shoot_max_temp_raster
```

### Precip cross-basis

Raster version:

```{r}
shoot_precip_raster <-
  ggplot(sp_pd,
         aes(x = x, y = lag, z = fitted, fill = fitted)) +
  geom_raster(interpolate = TRUE) +
  geom_raster(data = filter(sp_pd, min_dist >= 0.1), fill = "white", alpha = 0.4) +
  geom_contour(color = "grey50", size = 0.4, binwidth = contour_binwidth) +
  scale_fill_viridis_c("Shoot growth\n(cm/day)", option = "A", limits = growth_lims) + 
  scale_y_continuous("Lag (days)", expand = c(0,0)) + 
  scale_x_continuous("Daily Precipitation (mm)", expand = c(0,0))
shoot_precip_raster
```

Combine raster plots
```{r}
shoot_cb <-
  shoot_min_temp_raster + shoot_temp_raster + shoot_max_temp_raster + shoot_precip_raster +
  plot_layout(guides = "collect") &
  coord_cartesian(ylim = c(1,15)) &
  plot_annotation(tag_levels = "A") &
  theme_classic() &
  theme(text = element_text(size = 10))
```


```{r}
shoot_cb
ggsave(here("figs", "shoot_crossbasis.png"), shoot_cb, width = 165, height = 120, units = "mm")
ggsave(here("figs", "figure 3.eps"), shoot_cb, width = 165, height = 120, units = "mm")
```

Try adding a rug for weather data

```{r}
weather <- read_rds(here("data", "cleaned", "weather_lags.rds"))
weather <- weather %>% 
  filter(date > min(shootdf$date) - lubridate::days(15),
         date != max(shootdf$date)) #last day wouldn't show up in lags for any observations
range(weather$date)
range(shootdf$date)

shoot_cb_rug <- 
  shoot_min_temp_raster +
  geom_rug(data = weather, inherit.aes = FALSE, aes(x = temp_min),color = "white") +
shoot_temp_raster +
  geom_rug(data = weather, inherit.aes = FALSE, aes(x = temp_mean), color = "white") +
shoot_max_temp_raster +
  geom_rug(data = weather, inherit.aes = FALSE, aes(x = temp_max), color = "white") +
shoot_precip_raster +
  geom_rug(data = weather, inherit.aes = FALSE, aes(x = precip), color = "white")

shoot_cb_rug <-
  shoot_cb_rug +
  plot_layout(guides = "collect") &
  coord_cartesian(ylim = c(1,15)) &
  plot_annotation(tag_levels = "A") &
  theme_classic() &
  theme(text = element_text(size = 10))
```


```{r}
shoot_cb_rug
ggsave(here("figs", "shoot_crossbasis_rug.png"), shoot_cb_rug, width = 165, height = 120, units = "mm")
```
