---
title: "Leafhopper distributed lag model"
author: "Eric R. Scott"
date: '2020-01-25'
output:
  word_document:
    toc: yes
  html_notebook:
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
options(scipen = 4)
```

# TODO

-   for these new hypotheses tests to work, I need to fix the penalization for the covariates at whatever they are in the full model.

-   I might be able to do this with the `sp` parameter, or possibly by setting `k` and using `fx = TRUE`

*Last compiled: `r Sys.Date()`*

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(conflicted)
library(mgcv)
library(dlnm)
library(bbmle)

conflict_prefer("filter", "dplyr")
conflict_prefer("here", "here")
conflict_prefer("lag", "dplyr")
conflict_prefer("select", "dplyr")
```

# Methods

This analysis is based on Gasparrini A, Scheipl F, Armstrong B, Kenward MG (2017) A penalized framework for distributed lag non-linear models. Biometrics 73:938--948. doi: 10.1111/biom.12645. Specifically, it is based on `ex1_01.int.R` provided in their code. This "internal" method for distributed lag models takes full(?) advantage of built in features of `gam`s.

The goal is to model the effects of temperature and precipitation on leafhoppers as having a non-linear, lagged effect.

# Hypotheses

To achieve this goal, I'll test three hypotheses for each weather variable (min, mean, and max daily tempearture, and precipitation)

1.  Is *smooth* function necessary?

    -   H~A~: Full model including crossbasis smooth for weather variable over 15 days lag.
    -   H~0~: replace crossbasis term with mean of temperature in past 15 days

2.  Is there an effect of temperature?

    -   H~A~: Same as H~A~ from \#1 above
    -   H~0~: Remove the weather variable term entirely

3.  Is there a single lagged effect?

    -   H~A~: Use the weather variable at a lag with maximum effect suggested by the DLNM (e.g. mean temperature 10 days ago) as a parameteric (not smoothed) term.
    -   H~0~: Same as H~0~ from \#2 above

# Read in Data

```{r}
hoppers <- read_rds(here("data", "cleaned", "hopper-weather.rds"))

head(hoppers)
# str(hoppers)
```

Remove any rows with missing values

```{r}
hoppers <-
  hoppers %>%
  rowwise() %>%
  filter(!is.na(sum(Q_temp_max)))
```

```{r}
#calculate mean of past 15 days
hoppers <-
  hoppers %>%
  mutate(mean_temp_mean = mean(Q_temp_mean),
         mean_temp_min = mean(Q_temp_min),
         mean_temp_max = mean(Q_temp_max),
         mean_precip = mean(Q_precip)) %>% 
  #and the mean of days 9–12
  mutate(pick_temp_mean = mean(Q_temp_mean[ , 9:12]),
         pick_temp_min = mean(Q_temp_min[ , 9:12]),
         pick_temp_max = mean(Q_temp_max[ , 9:12]),
         pick_precip = mean(Q_temp_max[ , 9:12])) %>% 
  ungroup()
```

# Build the full models

Temp and precip are highly concurve, so cannot include all weather variables as predictors in the same model.

```{r}
m_hopper_temp_mean <- 
  gam(hoppers ~
        s(Q_temp_mean, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
        ) +
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")
```

```{r}
m_hopper_temp_min <- 
  gam(hoppers ~
        s(Q_temp_min, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
        ) +
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")
```

```{r}
m_hopper_temp_max <- 
  gam(hoppers ~
        s(Q_temp_max, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
        ) +
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")
```

```{r}
m_hopper_precip <- 
  gam(hoppers ~
        s(Q_precip, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
        ) +
        s(day_post, bs = "cr") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        # counter +
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")
```

```{r}
AICtab(m_hopper_temp_mean, m_hopper_temp_min, m_hopper_temp_max, m_hopper_precip)
```

Minimum temperature model "wins"

# Save models

```{r}
write_rds(m_hopper_precip, here("analysis", "models", "m_hopper_precip.rds"))
write_rds(m_hopper_temp_max, here("analysis", "models", "m_hopper_temp_max.rds"))
write_rds(m_hopper_temp_mean, here("analysis", "models", "m_hopper_temp_mean.rds"))
write_rds(m_hopper_temp_min, here("analysis", "models", "m_hopper_temp_min.rds"))
```

# Mean Temperature

## Inspect the GAM

```{r}
#check for overdispersion
sum(residuals(m_hopper_temp_mean, type = "pearson")^2) / df.residual(m_hopper_temp_mean)
gam.check(m_hopper_temp_mean)
```

Not overdispersed, appropriate numbers of knots used, residuals meet assumptions (mostly).

## Marginal Hypothesis test

P-values for covariates. P-values for crossbasis obtained through specific LRT below.

```{r}
anova.gam(m_hopper_temp_mean)
```

## H1: Is smooth necessary?

```{r}
sp_hopper_temp_mean <- m_hopper_temp_mean$sp

m_hopper_temp_mean1 <- 
  gam(hoppers ~
        mean_temp_mean +
        s(day_post, bs = "cr", sp = sp_hopper_temp_mean["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_mean["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_mean["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_mean["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h1_hopper_temp_mean <- anova.gam(m_hopper_temp_mean1, m_hopper_temp_mean,  test = "Chisq")
h1_hopper_temp_mean
```

Yes, smooth is significantly better than just a mean of the past 15 days.

## H2: Is there an effect of daily mean temp?

```{r}
m_hopper_temp_mean_null <- 
  gam(hoppers ~
        s(day_post, bs = "cr", sp = sp_hopper_temp_mean["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_mean["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_mean["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_mean["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h2_hopper_temp_mean <- anova.gam(m_hopper_temp_mean_null, m_hopper_temp_mean, test = "Chisq")
h2_hopper_temp_mean
#larger residual DF means smaller estimated degrees of freedom
```

Yes, there is a significant effect of lagged mean daily temperature on leafhoppers

## H3: Is there an effect of a specific lag?

According to plots, all models had a steep slope (strong effect) between 9 and 12 days of lag. I'll use the mean of those days as a predictor.

```{r}
m_hopper_temp_mean3 <- 
  gam(hoppers ~
       pick_temp_mean +
        s(day_post, bs = "cr", sp = sp_hopper_temp_mean["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_mean["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_mean["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_mean["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h3_hopper_temp_mean <- anova.gam(m_hopper_temp_mean_null, m_hopper_temp_mean3, test = "Chisq")
h3_hopper_temp_mean
```

Significant effect of mean temp 10 days ago

# Minimum temperature

## Inspect the GAM

```{r}
gam.check(m_hopper_temp_min)
#check for overdispersion
sum(residuals(m_hopper_temp_min, type = "pearson")^2) / df.residual(m_hopper_temp_min)
```

## Marginal Hypothesis test

P-values for covariates. P-values for crossbasis obtained through specific LRT below.

```{r}
anova.gam(m_hopper_temp_min)
```

## H1: Is smooth necessary?

```{r}
sp_hopper_temp_min <- m_hopper_temp_min$sp

m_hopper_temp_min1 <- 
  gam(hoppers ~
        mean_temp_min +
        s(day_post, bs = "cr", sp = sp_hopper_temp_min["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_min["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_min["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_min["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h1_hopper_temp_min <- anova.gam(m_hopper_temp_min1, m_hopper_temp_min,  test = "Chisq")
h1_hopper_temp_min
```

Yes, the smooth is a significant improvment over just mean minimum temperature

## H2: Is there an effect of min temp?

```{r}
m_hopper_temp_min_null <- 
  gam(hoppers ~
        s(day_post, bs = "cr", sp = sp_hopper_temp_min["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_min["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_min["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_min["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h2_hopper_temp_min <- anova.gam(m_hopper_temp_min_null, m_hopper_temp_min, test = "Chisq")
h2_hopper_temp_min
#larger residual DF means smaller estimated degrees of freedom
```

yes, highly significant effect of temperature

## H3: Is there an effect of a specific lag?

```{r}
m_hopper_temp_min3 <- 
  gam(hoppers ~
        pick_temp_min +
        s(day_post, bs = "cr", sp = sp_hopper_temp_min["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_min["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_min["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_min["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h3_hopper_temp_min <- anova.gam(m_hopper_temp_min_null, m_hopper_temp_min3, test = "Chisq")
h3_hopper_temp_min
```

# Maximum temperature

## Inspect the GAM

```{r}
gam.check(m_hopper_temp_max)
#check for overdispersion
sum(residuals(m_hopper_temp_max, type = "pearson")^2) / df.residual(m_hopper_temp_max)
```

## Marginal Hypothesis test

P-values for covariates. P-values for crossbasis obtained through specific LRT below.

```{r}
anova.gam(m_hopper_temp_max)
```

## H1: Is smooth necessary?

```{r}
sp_hopper_temp_max <- m_hopper_temp_max$sp
m_hopper_temp_max1 <- 
  gam(hoppers ~
        mean_temp_max +
        s(day_post, bs = "cr", sp = sp_hopper_temp_max["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_max["s(counter)"]) +
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_max["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_max["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h1_hopper_temp_max <- anova.gam(m_hopper_temp_max1, m_hopper_temp_max,  test = "Chisq")
h1_hopper_temp_max
```

Smooth is marginally significant

## H2: Is there an effect of max temp?

```{r}
m_hopper_temp_max_null <- 
  gam(hoppers ~
        s(day_post, bs = "cr", sp = sp_hopper_temp_max["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_max["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_max["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_max["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h2_hopper_temp_max <- anova.gam(m_hopper_temp_max_null, m_hopper_temp_max,
                           test = "Chisq")
h2_hopper_temp_max
#larger residual DF means smaller estimated degrees of freedom
```

Significant effect of crossbasis

## H3: Is there an effect of a specific lag?

```{r}
m_hopper_temp_max3 <- 
  gam(hoppers ~
        pick_temp_max +
        s(day_post, bs = "cr", sp = sp_hopper_temp_max["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_temp_max["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_temp_max["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_temp_max["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h3_hopper_temp_max <- anova.gam(m_hopper_temp_max_null, m_hopper_temp_max3, test = "Chisq")
h3_hopper_temp_max
```

# Precip

## Inspect GAM

```{r}
gam.check(m_hopper_precip)
sum(residuals(m_hopper_precip, type = "pearson")^2) / df.residual(m_hopper_precip)
```

## Marginal hypothesis test

```{r}
anova.gam(m_hopper_precip)
```

Marginally significant effect of day

## H1: Is smooth necessary?

```{r}
sp_hopper_precip <- m_hopper_precip$sp

m_hopper_precip1 <- 
  gam(hoppers ~
        mean_precip +
        s(day_post, bs = "cr", sp = sp_hopper_precip["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_precip["s(counter)"]) +
        s(interharvest_id, bs = "re",
          sp = sp_hopper_precip["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_precip["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h1_hopper_precip <- anova.gam(m_hopper_precip1, m_hopper_precip,  test = "Chisq")
h1_hopper_precip
```

Yes, crossbasis smooth is significantly better than just mean over past 15 days.

## H2: Is there an effect of max temp?

```{r}
m_hopper_precip_null <- 
  gam(hoppers ~
        s(day_post, bs = "cr", sp = sp_hopper_precip["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_precip["s(counter)"]) + 
        s(interharvest_id, bs = "re",
          sp = sp_hopper_precip["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_precip["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")
h2_hopper_precip <- anova.gam(m_hopper_precip_null, m_hopper_precip, test = "Chisq")
h2_hopper_precip
#larger residual DF means smaller estimated degrees of freedom
```

Precipitation has a significant effect.

## H3: Is there an effect of a specific lag?

```{r}
m_hopper_precip3 <- 
  gam(hoppers ~
        pick_precip +
        s(day_post, bs = "cr", sp = sp_hopper_precip["s(day_post)"]) +
        s(counter, bs = "re", sp = sp_hopper_precip["s(counter)"]) +
        s(interharvest_id, bs = "re", 
          sp = sp_hopper_precip["s(interharvest_id)"]) +
        s(plant, bs = "re", sp = sp_hopper_precip["s(plant)"]),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2,
      method = "REML")

h3_hopper_precip <- anova.gam(m_hopper_precip_null, m_hopper_precip3, test = "Chisq")
h3_hopper_precip
```

# Table

Collect model statistics:

```{r}
full_models <- list(m_hopper_temp_min,
                    m_hopper_temp_mean,
                    m_hopper_temp_max,
                    m_hopper_precip)
h1 <- list(h1_hopper_temp_min,
           h1_hopper_temp_mean,
           h1_hopper_temp_max,
           h1_hopper_precip)

h2 <- list(h2_hopper_temp_min,
           h2_hopper_temp_mean,
           h2_hopper_temp_max,
           h2_hopper_precip)
h3 <- list(h3_hopper_temp_min,
           h3_hopper_temp_mean,
           h3_hopper_temp_max,
           h3_hopper_precip)

df <- 
  tibble(weather_var = c("Minimum daily temp.",
                         "Mean daily temp.",
                         "Maximum daily temp.",
                         "Total daily precip."),
         r2 = map_dbl(full_models, ~summary(.x)$r.sq),
         aic = map_dbl(full_models, AIC),
         h1_df = map_dbl(h1, ~pull(.x, "Df")[2]),
         h1_x2 = map_dbl(h1, ~pull(.x, "Deviance")[2]),
         h1_p = map_dbl(h1, ~pull(.x, "Pr(>Chi)")[2]),
         h2_df = map_dbl(h2, ~pull(.x, "Df")[2]),
         h2_x2 = map_dbl(h2, ~pull(.x, "Deviance")[2]),
         h2_p = map_dbl(h2, ~pull(.x, "Pr(>Chi)")[2]),
         h3_df = map_dbl(h3, ~pull(.x, "Df")[2]),
         h3_x2 = map_dbl(h3, ~pull(.x, "Deviance")[2]),
         h3_p = map_dbl(h3, ~pull(.x, "Pr(>Chi)")[2]),
  ) 

df

```


```{r}
library(flextable)
get_flextable_defaults()
set_flextable_defaults(font.family = "Times New Roman",
                       font.size = 9)
```


```{r}
df %>% 
  mutate(across(ends_with("_p"), ~gtsummary::style_pvalue(.x))) %>% 
  flextable(col_keys = c("weather_var", "r2", "aic", "sep_1", "h1_df", "h1_x2", "h1_p", "sep_2", "h2_df", "h2_x2", "h2_p", "sep_3", "h3_df", "h3_x2", "h3_p")) %>% 
  colformat_num(j = 3, big.mark = "") %>% 
  colformat_num(j = c(4, 7, 10), digits = 3) %>% 
  colformat_num(j = c(5, 8, 11), digits = 4) %>% 
  add_header_row(c("", "DLNM", "", "H1", "", "H2", "", "H3"), colwidths = c(1, 2, 1, 3, 1, 3, 1, 3), top = TRUE) %>% 
  set_header_labels(
    weather_var = "",
    r2 = "R^2^",
    aic = "AIC",
    h1_df = "df",
    h1_x2 = "X^2^",
    h1_p = "P",
    h2_df = "df",
    h2_x2 = "X^2^",
    h2_p = "P",
    h3_df = "df",
    h3_x2 = "X^2^",
    h3_p = "P"
  ) %>% 
  merge_v(part = "header") %>% 
  autofit() %>%
  theme_booktabs() %>% 
  align(align = "center", part =  "header") %>% 
  align(i = 2, align = "right", part = "header") %>% 
  align(align = "right", part = "body")
```

# Exploration of how edf works

If I constrain the `sp` parameter, does it constrain the resulting EDF? Mostly...

```{r}
library(magrittr)
m_hopper_temp_mean$edf %>%
  enframe() %>%
  mutate(name = str_remove(name, ".\\d+")) %>%
  group_by(name) %>%
  summarize(edf = sum(value)) %T>%
  print() %>% 
  summarize(total = sum(edf))

m_hopper_temp_mean1$edf %>%
  enframe() %>%
  mutate(name = str_remove(name, ".\\d+")) %>%
  group_by(name) %>%
  summarize(edf = sum(value)) %T>%
  print() %>% 
  summarize(total = sum(edf))

m_hopper_temp_mean_null$edf %>%
  enframe() %>%
  mutate(name = str_remove(name, ".\\d+")) %>%
  group_by(name) %>%
  summarize(edf = sum(value)) %T>%
  print() %>% 
  summarize(total = sum(edf))

m_hopper_temp_mean3$edf %>%
  enframe() %>%
  mutate(name = str_remove(name, ".\\d+")) %>%
  group_by(name) %>%
  summarize(edf = sum(value)) %T>%
  print() %>% 
  summarize(total = sum(edf))
```
