---
title: "Leafhopper distributed lag model"
author: "Eric R. Scott"
date: "2020-01-25"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---
#TODO

- Tidy up analysis
- export figures
- make sure doing separate GAMs for temp and precip is legit


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Last compiled: `r Sys.Date()`*

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(lme4)
library(car)
library(broom)
library(plotly)
library(here)
library(conflicted)
library(bbmle)
library(lmtest)
library(mgcv)
library(dlnm)
library(MASS)

conflict_prefer("filter", "dplyr")
conflict_prefer("here", "here")
conflict_prefer("lag", "dplyr")
conflict_prefer("select", "dplyr")
```

# Goal

This analysis is based on Gasparrini A, Scheipl F, Armstrong B, Kenward MG (2017) A penalized framework for distributed lag non-linear models. Biometrics 73:938–948. doi: 10.1111/biom.12645.  Specifically, it is based on `ex1_01.int.R` provided in their code.  This "internal" method for distributed lag models takes full(?) advantage of built in features of `gam`s.

The goal is to model the effects of temperature and precipitation on leafhoppers as having a quadratic, lagged effect.  The lag is modeled as a penalized cubic regression spline and the effect of temperature as a 2nd order polynomial in a crossbasis smoother. In an ideal world, the penalized spline would be the shrinkage version (bs = "cs"), but that isn't available in the crossbasis smooth constructor provided by `dlnm`.

# Read in Data

```{r}
hoppers <- read_rds(here("data", "cleaned", "hopper-weather.rds"))

hoppers <- hoppers %>% 
  mutate(interharvest_id = as.factor(paste0(field, interharvest.no)),
         counter = as.factor(counter),
         plant = as.factor(plant)) %>% 
  filter(!is.na(precip_lag14)) #get rid of plants that don't have 14 days of data

head(hoppers)
# str(hoppers)
```
```{r}
# plot(hoppers$temp_mean, hoppers$precip_mm)
# cor.test(hoppers$temp_mean, hoppers$precip_mm)
# 
# plot(hoppers$counter, hoppers$temp_mean)
```

# Simple glm() for sanity check

```{r}
m1 <- glmer(hoppers ~ scale(temp_lag2) + scale(precip_lag2) + day_post + counter + (1|interharvest_id),
          offset = log(n_leaves), family = poisson(link = "log"), data = hoppers)
summary(m1)
Anova(m1)

#check for overdispersion.  Should be about 1
sum(residuals(m1, type = "pearson")^2) / df.residual(m1)
```

'counter' is highly significant, meaning that the three of us had different detection probabilities.

```{r}
m1.a <- glmer(hoppers ~ -1 + scale(temp_lag2) + scale(precip_lag2) + day_post + counter + (1|interharvest_id),
          offset = log(n_leaves), family = poisson(link = "log"), data = hoppers)
fixef(m1.a)[4:6] %>% exp()
```

Can (should) I use `counter` as a random effect with only three levels?

```{r}
m1.b <- glmer(hoppers ~  scale(temp_lag2) + scale(precip_lag2) + day_post + (1|counter) + (1|interharvest_id),
          offset = log(n_leaves), family = poisson(link = "log"), data = hoppers)
summary(m1.b)
Anova(m1.b)
```

I get similar estimates for coefficients, so I'm going to say yes???

## with glm.nb

```{r}
m1.nb <- glmer.nb(hoppers ~ scale(temp_lag2) + scale(precip_lag2) + day_post + (1|counter) + (1|interharvest_id),
          offset = log(n_leaves), data = hoppers)
summary(m1.nb)
Anova(m1.nb)

#check for overdispersion.  Should be about 1
sum(residuals(m1.nb, type = "pearson")^2) / df.residual(m1.nb)
```

Dispersion is closer to 1, but model fails to converge.  Might not be a problem in a GAM though.


# Distributed Lag Model: "internal" method

## Setup

Create matrices for temperature, precip, and lag.

```{r}
Q_temp <-
  hoppers %>% 
  select(starts_with("temp"), -temp_at_count, -temp_max, -temp_min) %>% 
  as.matrix()

Q_precip <-
  hoppers %>%
  select(starts_with("precip")) %>% 
  as.matrix()

L <- matrix(0:(ncol(Q_temp)-1), nrow(Q_temp), ncol(Q_temp), byrow = TRUE)
```

# Build the GAMs

Seems like if I use the offset it makes things *worse*.  Like, way overdispersed with poisson, but underdispersed with negbin.  Also, gam.check() just crashes if I use offset and family = nb at the same time

```{r}
m_temp <- 
  gam(hoppers ~
        s(Q_temp, L,
          bs = "cb",
          k = c(14, 14),
          xt = list(bs = "cr")
        ) +
        # s(Q_precip, L,
        #   bs = "cb",
        #   k = c(14, 14),
        #   xt = list(bs = "cr")
        # ) +
        s(day, bs = "cs") +
        s(day_post, bs = "cs") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2, #values above 1 make smoother.  Borrowed from Teller et al.
      method = "REML")

m_precip <- 
  gam(hoppers ~
        # s(Q_temp, L,
        #   bs = "cb",
        #   k = c(14, 14),
        #   xt = list(bs = "cr")
        # ) +
        s(Q_precip, L,
          bs = "cb",
          k = c(14, 14),
          xt = list(bs = "cr")
        ) +
        s(day, bs = "cs") +
        s(day_post, bs = "cs") +
        s(counter, bs = "re") + #no difference with random vs. fixed
        s(interharvest_id, bs = "re") +
        s(plant, bs = "re"),
      family = poisson(),
      offset = log(n_leaves),
      data = hoppers,
      gamma = 1.2, #values above 1 make smoother.  Borrowed from Teller et al.
      method = "REML")
```
```{r}
AIC(m_temp, m_precip)
```

AICs nearly identical!  Keep both models??

# Temperature

## Inspect the GAM

```{r}
gam.check(m_temp)
sum(residuals(m_temp, type = "pearson")^2) / df.residual(m_temp)
```

not overdispersed

```{r}
concurvity(m_temp, full = TRUE)
# concurvity(m_temp, full = FALSE)
```

I'm a little bit confused how temperature and counter could have such high concurvity.  Might want to check for systematic changes in counter.  I guess wei ji peng wasn't there the entire time.  Maybe he was gone for the hottest week?? 

## Significance test

```{r}
anova.gam(m_temp)
```
Significant effect of plant (parametric factor), temperature, precip, days post harvest, and diameter on growth.

## Predictions and plots

### Plot regular smooths

```{r}
plot(m_hopper, pages = 1, residuals = FALSE)
```



### Plot crossbasis smooths

```{r}
pred_temp <- crosspred("Q_temp", m_temp)
pred_temp2 <- crosspred("Q_temp", m_temp, by = 0.2, bylag = 0.2, cen = 28)

#3D
plot(pred_temp,
     ptype = "contour",
     xlab = "Temperature (C)",
     ylab = "lag (days)",
     main = "Temperature")

# Overall effect of temp
plot(pred_temp2,
     "overall",
     # ci = "bars",
     ylab = "Relative change in leafhoppers / leaf",
     xlab = "Temperature (C)",
     main = "Temperature Effects")

plot(pred_temp2,
     var = 21,
     xlab = "Lag (days)",
     ylab = "Relative change in leafhoppers / leaf",
     main = "Effect of Lag for 21ºC")
```

Significant effect of lagged temperature, but not of date.  The further back, the greater the effect (means weather data doesnt' go far enough back to capture effect?).  Cool temperatures increase leafhopper density?!??!??!?!?!?!?!?!?!?!?

# Precip
```{r}
gam.check(m_precip)
```

```{r}
anova.gam(m_precip)
```

significant effect of precip, but not of date.

```{r}
pred_precip <- crosspred("Q_precip", m_precip, by = 0.05,
                         # cen = 32
                         )
pred_precip2 <- crosspred("Q_precip", m_precip, by = 0.2, bylag = 0.2)

#3D
plot(pred_precip,
     ptype = "contour",
     xlab = "Precip (mm)",
     ylab = "lag (days)",
     main = "Precip")

# Overall effect of precip
plot(pred_precip2,
     "overall",
     ylab = "Relative Growth Rate",
     xlab = "Precip (mm)",
     main = "Precip")

plot(pred_precip2,
     var = 0,
     xlab = "Lag (days)",
     ylab = "Relative Growth Rate",
     main = "Effect of Lag for 0 mm")

plot(pred_precip2,
     var = 80,
     xlab = "Lag (days)",
     ylab = "Relative Growth Rate",
     main = "Effect of Lag for 80 mm")
#CI overlaps 1 for entire range
```

Low precip at 5 and 11 days ago has negative effect on leafhoppers (dessication?)

High precip 8 days ago has positive effect on leafhoppers

# Conclusion