---
title: "Analysis Title"
author: "Eric R. Scott"
date: "Date Initiated"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

#TODO

3. Include plant as a random effect rather than a fixed effect.  Probably will require use of `gamm4` package.
Should be re-coded as just a number and nested inside `harvest_id`.  Like: `+ (1|harvest_id) + (harvest_id|plant)`, I think.

include conversion to factors in wrangling, not in this doc.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Last compiled: `r Sys.Date()`*

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(lme4)
library(car)
library(broom)
library(plotly)
library(here)
library(conflicted)
library(bbmle)
library(lmtest)
library(mgcv)
library(dlnm)

conflict_prefer("filter", "dplyr")
conflict_prefer("here", "here")
conflict_prefer("lag", "dplyr")
conflict_prefer("select", "dplyr")
```

# Goal

This analysis is based on Gasparrini A, Scheipl F, Armstrong B, Kenward MG (2017) A penalized framework for distributed lag non-linear models. Biometrics 73:938–948. doi: 10.1111/biom.12645.  Specifically, it is based on `ex1_01.int.R` provided in their code.  This "internal" method for distributed lag models takes full(?) advantage of built in features of `gam`s.

The goal is to model the effects of temperature and precipitation on plant growth as having a quadratic, lagged effect.  The lag is modeled as a penalized cubic regression spline and the effect of temperature as a 2nd order polynomial in a crossbasis smoother. In an ideal world, the penalized spline would be the shrinkage version (bs = "cs"), but that isn't available in the crossbasis smooth constructor provided by `dlnm`.

# Read in Data

```{r}
shootdf <- read_rds(here("data", "cleaned", "shoots.rds"))
head(shootdf)
shootdf <- shootdf %>% 
  separate(shoot_id, into = c("field", "harvest", "plant", "shoot"), remove = FALSE, sep = "-") %>% 
  filter(!is.na(precip_lag14)) #get rid of plants that don't have 7 days of data
```

# Distributed Lag Model: "internal" method

## Setup

Create matrices for temperature, precip, and lag.

```{r}
temp_mat <-
  shootdf %>% 
  select(starts_with("temp")) %>% 
  as.matrix()

precip_mat <-
  shootdf %>%
  select(starts_with("precip")) %>% 
  as.matrix()

# Q_temp <- temp_mat[ , 1:6]
Q_temp <- temp_mat #use all the data
# L <- matrix(0:5, nrow(Q_temp), ncol(Q_temp), byrow = TRUE)

L <- matrix(0:(ncol(Q_temp) - 1), nrow(Q_temp), ncol(Q_temp), byrow = TRUE)

# Q_precip <- precip_mat[ , 1:6]
Q_precip <- precip_mat
```


Create options that override first dimension of smoother with a polynomial function instead.

```{r}
poly <- dlnm:::poly
xt <- list(bs = "cr", argvar = list(fun = "poly", degree = 2))
```


## Build the GAM

```{r}
shootdf <- shootdf %>% mutate(plant = as.factor(plant), interharvest_id = as.factor(interharvest_id))
w_gam_int <- 
  gam(growth ~
        s(Q_temp, L,
          bs = "cb",
          k = c(14, 14),
          # xt = xt
          xt = list(bs = "cr")
          ) +
        # s(Q_precip, L,
        #   bs = "cb",
        #   k = c(7, 7),
        #   xt = xt
        #   ) +
        s(day_post, bs = "cs", k = 15) + s(diameter, bs = "cs") + #fixed effects
        s(interharvest_id, bs = "re") + s(plant, bs = "re"), #random effects
      data = shootdf,
      family = scat,
      gamma = 1.2, #values above 1 make smoother.  Borrowed from Teller et al.
      method = "REML")
```
 

## Inspect the GAM

```{r}
gam.check(w_gam_int)
```

GAM converges, day_post had low p-value for the basis dimension check, so increased k to 15, but stopped there because it didn't increase edf at all.

```{r}
concurvity(w_gam_int, full = FALSE)
```

Both precip and temperature have potential problems with concurvity

Precip and temp are concurve(?) meaning they might be redundant in the model.  This makes sense as temperature and precip are generally correlated to some extent.

## Significance test

```{r}
anova.gam(w_gam_int)
summary(w_gam_int)$edf
w_gam_int$sp
```
Significant effect of plant (parametric factor), temperature, precip, days post harvest, and diameter on growth.

## Predictions and plots

### Plot regular smooths

```{r}
plot(w_gam_int, pages = 1, residuals = FALSE)
```
Shoots with larger diameters grow faster.  This looks like it could be a linear term, even though the edf > 1.  Shoot growth rates vary through time in a more complex manner with the fastest growth happening about 10 days post harvest and the slowest growth around 25 days post harvest.  The uptick at the end must be spurrious.


### Plot crossbasis smooths

```{r}
pred_temp <- crosspred("Q_temp", w_gam_int, cen = 22)
pred_temp2 <- crosspred("Q_temp", w_gam_int, by = 0.2, bylag = 0.2, cen = 22)

#3D
plot(pred_temp,
     ptype = "contour",
     xlab = "Temperature (C)",
     ylab = "lag (days)",
     main = "Temperature")

# Overall effect of temp
plot(pred_temp2,
     "overall",
     ylab = "Relative Growth Rate",
     xlab = "Temperature (C)",
     main = "Temperature Effects")

plot(pred_temp2,
     var = 26,
     xlab = "Lag (days)",
     ylab = "Relative Growth Rate",
     main = "Effect of Lag for 26ºC")
```



```{r}
pred_precip <- crosspred("Q_precip", w_gam_int, cen = 63, by = 0.05)
pred_precip2 <- crosspred("Q_precip", w_gam_int, by = 0.2, bylag = 0.2, cen = 63)

#3D
plot(pred_precip,
     ptype = "contour",
     xlab = "Precip (mm)",
     ylab = "lag (days)",
     main = "Precip")

# Overall effect of temp
plot(pred_precip2,
     "overall",
     ylab = "Relative Growth Rate",
     xlab = "Precip (mm)",
     main = "Precip")

plot(pred_precip2,
     var = 20,
     xlab = "Lag (days)",
     ylab = "Relative Growth Rate",
     main = "Effect of Lag for 20 mm")
```

Strong positive effect of low precipitation 0-2 days ago. Negative effect of very high precip. This makes sense as low precip is probably associated with more sun and higher temperatures.

# Conclusion