---
title: "Distributed Lag Model for Shoot Growth"
author: "Eric R. Scott"
date: "2020-01-15"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print=FALSE)
options(scipen = 4)
```

*Last compiled: `r Sys.Date()`*

# Packages

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(conflicted)
library(mgcv)
library(dlnm)
library(bbmle)

conflict_prefer("filter", "dplyr")
conflict_prefer("here", "here")
conflict_prefer("lag", "dplyr")
conflict_prefer("select", "dplyr")

source(here("R", "crossbasis.r"))
```

# Goal

This analysis is based on Gasparrini A, Scheipl F, Armstrong B, Kenward MG (2017) A penalized framework for distributed lag non-linear models. Biometrics 73:938–948. doi: 10.1111/biom.12645.  Specifically, it is based on `ex1_01.int.R` provided in their code.  This "internal" method for distributed lag models takes full(?) advantage of built in features of `gam`s.

The goal is to model the effects of temperature and precipitation on plant growth allowing for variable shape through time (lag) and across values of the predictor using a two-dimensional smoothing function called a cross-basis function. In an ideal world, the penalized spline would be the shrinkage version (bs = "cs"), but that isn't available in the crossbasis smooth constructor provided by `dlnm`, so this uses regular penalized cubic regression splines.

# Read in Data

```{r}
shootdf <- read_rds(here("data", "cleaned", "shoots.rds"))
```

Variables I'll be using:

- `growth` (response): a linear growth rate in cm ($h_t - h_{t-1}$)
- `diameter` (co-variate): stem diameter in mm
- `day_post` (co-variate): number of days post-harvest.  Shoots are "re-set" every harvest
- `temp_*` (predictor): I'm using mean temperature lagged out to 14 days to make a cross-basis function
- `precip_*` (predictor): precipitation lagged out to 14 days
- `plant` (random effect): which plant was the shoot on
- `interharvest_id` (random effect): combination of field ID and which interharvest period it was.  Each field x interharvest period experienced a unique climate history because harvests were not synchronized between the two fields.

```{r}
#calculate mean of past 15 days
shootdf <-
  shootdf %>%
  rowwise() %>% 
  mutate(mean_temp_mean = mean(Q_temp_mean),
         mean_temp_min = mean(Q_temp_min),
         mean_temp_max = mean(Q_temp_max),
         mean_precip = mean(Q_precip)) %>% 
  #and the mean of days 1--4
  mutate(pick_temp_mean = mean(Q_temp_mean[ , 1:4]),
         pick_temp_min =   mean(Q_temp_min[ , 1:4]),
         pick_temp_max =   mean(Q_temp_max[ , 1:4]),
         pick_precip =     mean(Q_temp_max[ , 1:4])) %>% 
  ungroup()
```

# Build full models

```{r}
m_shoot_temp_mean <- 
  gam(growth ~ 
        #cross-basis function for temperature
        s(Q_temp_mean, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
          ) +
        s(day_post, bs = "cr") + s(diameter, k = 15, bs = "cr") + #co-variates
        s(interharvest_id, bs = "re") + s(plant, bs = "re"), #random effects
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
```


```{r}
m_shoot_temp_min <- 
  gam(growth ~ 
        #cross-basis function for temperature
        s(Q_temp_min, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
          ) +
        s(day_post, bs = "cr") + s(diameter, k = 15, bs = "cr") + #co-variates
        s(interharvest_id, bs = "re") + s(plant, bs = "re"), #random effects
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
```


```{r}
m_shoot_temp_max <- 
  gam(growth ~ 
        #cross-basis function for temperature
        s(Q_temp_max, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
          ) +
        s(day_post, bs = "cr") + s(diameter, k = 15, bs = "cr") + #co-variates
        s(interharvest_id, bs = "re") + s(plant, bs = "re"), #random effects
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
```

```{r}
m_shoot_precip <- 
  gam(growth ~ 
        #cross-basis function for precipitation
        s(Q_precip, L,
          bs = "cb",
          k = c(10, 14),
          xt = list(bs = "cr")
          ) +
        s(day_post, bs = "cr") + s(diameter, k = 15, bs = "cr") + #co-variates
        s(interharvest_id, bs = "re") + s(plant, bs = "re"), #random effects
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
```


```{r}
ICtab(m_shoot_temp_mean, m_shoot_temp_min, m_shoot_temp_max, m_shoot_precip)
```

temperature models have the lowest AIC.  Min, max, mean all essentially interchangable.

# Save Models

```{r}
write_rds(m_shoot_precip, here("analysis", "models",    "m_shoot_precip.rds"))
write_rds(m_shoot_temp_max, here("analysis", "models",  "m_shoot_temp_max.rds"))
write_rds(m_shoot_temp_mean, here("analysis", "models", "m_shoot_temp_mean.rds"))
write_rds(m_shoot_temp_min, here("analysis", "models",  "m_shoot_temp_min.rds"))
```



# Mean Temperature

## Inspect the GAM

```{r}
gam.check(m_shoot_temp_mean)
```
basis dimensions seem fine despite low p-value for k-index.

## Marginal hypothesis test

```{r}
anova(m_shoot_temp_mean)
```

Significant effects of lagged temperature, days post harvest, stem diameter, and plant ID.

## H1: Is smooth necessary?

```{r}
sp_shoot_temp_mean <- m_shoot_temp_mean$sp

m_shoot_temp_mean1 <- 
  gam(growth ~ 
        mean_temp_mean +
        s(day_post, bs = "cr", sp = sp_shoot_temp_mean["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_mean["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_mean["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_mean["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
h1_shoot_temp_mean <- anova.gam(m_shoot_temp_mean1, m_shoot_temp_mean,  test = "Chisq")
h1_shoot_temp_mean
```

## H2: Is there an effect of crossbasis function?

```{r}
m_shoot_temp_mean_null <- 
  gam(growth ~ 
        s(day_post, bs = "cr", sp = sp_shoot_temp_mean["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_mean["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_mean["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_mean["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")

h2_shoot_temp_mean <- anova.gam(m_shoot_temp_mean_null, m_shoot_temp_mean, test = "Chisq")
h2_shoot_temp_mean
#larger residual DF means smaller estimated degrees of freedom
```

## H3: Is there an effect of a specific lag time?

```{r}
m_shoot_temp_mean3 <- 
  gam(growth ~ 
        pick_temp_mean +
        s(day_post, bs = "cr", sp = sp_shoot_temp_mean["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_mean["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_mean["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_mean["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
h3_shoot_temp_mean <- anova.gam(m_shoot_temp_mean_null, m_shoot_temp_mean3, test = "Chisq")
h3_shoot_temp_mean
```

# Minimum Temperature

## Inspect the GAM

```{r}
gam.check(m_shoot_temp_min)
```

## Marginal Hypothesis test

```{r}
anova(m_shoot_temp_min)
```

## H1: Is smooth necessary?

```{r}
sp_shoot_temp_min <- m_shoot_temp_min$sp

m_shoot_temp_min1 <- 
  gam(growth ~ 
        mean_temp_min +
        s(day_post, bs = "cr", sp = sp_shoot_temp_min["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_min["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_min["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_min["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
h1_shoot_temp_min <- anova.gam(m_shoot_temp_min1, m_shoot_temp_min,  test = "Chisq")
h1_shoot_temp_min
```

## H2: Is there an effect of crossbasis function?

```{r}
m_shoot_temp_min_null <- 
  gam(growth ~ 
        s(day_post, bs = "cr", sp = sp_shoot_temp_min["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_min["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_min["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_min["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")

h2_shoot_temp_min <- anova.gam(m_shoot_temp_min_null, m_shoot_temp_min, test = "Chisq")
h2_shoot_temp_min
#larger residual DF means smaller estimated degrees of freedom
```

## H3: Is there an effect of a specific lag time?

```{r}
m_shoot_temp_min3 <- 
  gam(growth ~ 
        pick_temp_min +
        s(day_post, bs = "cr", sp = sp_shoot_temp_min["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_min["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_min["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_min["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
h3_shoot_temp_min <- anova.gam(m_shoot_temp_min_null, m_shoot_temp_min3, test = "Chisq")
h3_shoot_temp_min
```
# Maximum temperature

## Inspect the GAM

```{r}
gam.check(m_shoot_temp_max)
```
basis dimensions seem fine despite low p-value for k-index.

## Marginal hypothesis test

```{r}
anova(m_shoot_temp_max)
```

Significant effects of lagged temperature, days post harvest, stem diameter, and plant ID.

## H1: Is smooth necessary?

```{r}
sp_shoot_temp_max <- m_shoot_temp_max$sp

m_shoot_temp_max1 <- 
  gam(growth ~ 
        mean_temp_max +
        s(day_post, bs = "cr", sp = sp_shoot_temp_max["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_max["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_max["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_max["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
h1_shoot_temp_max <- anova.gam(m_shoot_temp_max1, m_shoot_temp_max,  test = "Chisq")
h1_shoot_temp_max
```

## H2: Is there an effect of crossbasis function?

```{r}
m_shoot_temp_max_null <- 
  gam(growth ~ 
        s(day_post, bs = "cr", sp = sp_shoot_temp_max["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_max["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_max["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_max["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")

h2_shoot_temp_max <- anova.gam(m_shoot_temp_max_null, m_shoot_temp_max, test = "Chisq")
h2_shoot_temp_max
#larger residual DF means smaller estimated degrees of freedom
```

## H3: Is there an effect of a specific lag time?

```{r}
m_shoot_temp_max3 <- 
  gam(growth ~ 
        pick_temp_max +
        s(day_post, bs = "cr", sp = sp_shoot_temp_max["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_temp_max["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_temp_max["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_temp_max["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
h3_shoot_temp_max <- anova.gam(m_shoot_temp_max_null, m_shoot_temp_max3, test = "Chisq")
h3_shoot_temp_max
```
# Precipitation model

## Inspect the GAM

```{r}
gam.check(m_shoot_precip)
```
basis dimensions seem fine despite low p-value for k-index.

## Marginal hypothesis test

```{r}
anova(m_shoot_precip)
```

Interstingly, here the days post harvest term is not significant.

## H1: Is smooth necessary?

```{r}
sp_shoot_precip <- m_shoot_precip$sp

m_shoot_precip1 <- 
  gam(growth ~ 
        mean_precip +
        s(day_post, bs = "cr", sp = sp_shoot_precip["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_precip["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_precip["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_precip["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
h1_shoot_precip <- anova.gam(m_shoot_precip1, m_shoot_precip,  test = "Chisq")
h1_shoot_precip
```

## H2: Is there an effect of crossbasis function?

```{r}
m_shoot_precip_null <- 
  gam(growth ~ 
        s(day_post, bs = "cr", sp = sp_shoot_precip["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_precip["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_precip["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_precip["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")

h2_shoot_precip <- anova.gam(m_shoot_precip_null, m_shoot_precip, test = "Chisq")
h2_shoot_precip
#larger residual DF means smaller estimated degrees of freedom
```

## H3: Is there an effect of a specific lag time?

```{r}
m_shoot_precip3 <- 
  gam(growth ~ 
        pick_precip +
        s(day_post, bs = "cr", sp = sp_shoot_precip["s(day_post)"]) + 
        s(diameter, k = 15, bs = "cr", sp = sp_shoot_precip["s(diameter)"]) + 
        s(interharvest_id, bs = "re", sp = sp_shoot_precip["s(interharvest_id)"]) + 
        s(plant, bs = "re", sp = sp_shoot_precip["s(plant)"]), 
      data = shootdf,
      family = scat, #because residuals are very leptokurtic w/ gaussian
      gamma = 1.2,
      method = "REML")
h3_shoot_precip <- anova.gam(m_shoot_precip_null, m_shoot_precip3, test = "Chisq")
h3_shoot_precip
```

# Table

Collect model statistics:

```{r}
full_models <- list(m_shoot_temp_min,
                    m_shoot_temp_mean,
                    m_shoot_temp_max,
                    m_shoot_precip)
h1 <- list(h1_shoot_temp_min,
           h1_shoot_temp_mean,
           h1_shoot_temp_max,
           h1_shoot_precip)

h2 <- list(h2_shoot_temp_min,
           h2_shoot_temp_mean,
           h2_shoot_temp_max,
           h2_shoot_precip)

h3 <- list(h3_shoot_temp_min,
           h3_shoot_temp_mean,
           h3_shoot_temp_max,
           h3_shoot_precip)

df <- 
  tibble(weather_var = c("Minimum daily temperature",
                         "Mean daily temperature",
                         "Maximum daily temperature",
                         "Total daily precipitation"),
         r2 = map_dbl(full_models, ~summary(.x)$r.sq),
         aic = map_dbl(full_models, AIC),
         h1_df = map_dbl(h1, ~pull(.x, "Df")[2]),
         h1_x2 = map_dbl(h1, ~pull(.x, "Deviance")[2]),
         h1_p = map_dbl(h1, ~pull(.x, "Pr(>Chi)")[2]),
         h2_df = map_dbl(h2, ~pull(.x, "Df")[2]),
         h2_x2 = map_dbl(h2, ~pull(.x, "Deviance")[2]),
         h2_p = map_dbl(h2, ~pull(.x, "Pr(>Chi)")[2]),
         h3_df = map_dbl(h3, ~pull(.x, "Df")[2]),
         h3_x2 = map_dbl(h3, ~pull(.x, "Deviance")[2]),
         h3_p = map_dbl(h3, ~pull(.x, "Pr(>Chi)")[2]),
  ) 

df

```


```{r}
library(gt)
library(gtsummary)
# gt(df)
```

Yeah, seems like I can copy-paste into Word just fine.

```{r}
table2 <- 
  gt(df, rowname_col = "weather_var") %>% 
  tab_spanner(
    label = "DLNM",
    columns = vars(r2, aic)
  ) %>% 
  tab_spanner(
    label = md("H1"),
    columns = starts_with("h1_")
  ) %>% 
  tab_spanner(
    label = "H2",
    columns = starts_with("h2")
  ) %>% 
  tab_spanner(
    label = "H3 (mean of lags 1–4)",
    columns = starts_with("h3")
  )

table2 %>% 
  fmt(
    columns = ends_with("_p"),
    fns = gtsummary::style_pvalue
  ) %>% 
  fmt_number(
    columns = vars(r2, h1_df, h1_x2, h2_df, h2_x2, h3_df, h3_x2),
    decimals = 2
  ) %>% 
  fmt(
    columns = vars(aic),
    fns = round
  ) %>% 
  cols_label(
    r2 = html("R<sup>2</sup>"),
    aic = "AIC",
    h1_df = "df",
    h2_df = "df",
    h3_df = "df",
    # h1_x2 = md("$\\Chi^2$") #markdown not working?
    h1_x2 = html("X<sup>2</sup>"),
    h2_x2 = html("X<sup>2</sup>"),
    h3_x2 = html("X<sup>2</sup>"),
    h1_p = "P",
    h2_p = "P",
    h3_p = "P"
  )
```
