---
title: "Analysis Title"
author: "Eric R. Scott"
date: "Date Initiated"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

#TODO

Consider adding penalty to lag.  The further back in time, the more likely the effects are to be spurrious for this analysis (in fact, you probably dont' need this fancy analysis for shoot growth).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Last compiled: `r Sys.Date()`*

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(lme4)
library(car)
library(broom)
library(plotly)
library(here)
library(conflicted)
library(bbmle)
library(lmtest)
conflict_prefer("filter", "dplyr")
conflict_prefer("here", "here")
conflict_prefer("lag", "dplyr")
```

# Read in Data

```{r}
shootdf <- read_rds(here("data", "cleaned", "shoots.rds"))
head(shootdf)
```


# Distributed Lag Model

```{r}
library(dlnm)
library(mgcv)
```

## Construct lagged variables

Because or the complex design, I need to supply `crossbasis` with a matrix of pre-constructed lagged temperature variables. 


```{r}
temp_mat <-
  shootdf %>%
  ungroup() %>%
  select(starts_with("temp")) %>% 
  as.matrix()

precip_mat <-
  shootdf %>%
  ungroup() %>%
  select(starts_with("precip")) %>% 
  as.matrix()
```

Then I construct the cross-basis variable using a penalized cubic spline for the lag kernel and either a polynomial or linear relationship to the response variable.  For the cubic spline, I reduced the degrees of freedom to 6 because it was causing problems with `lmer` with the default.

```{r}
temp_k <-
  crossbasis(
  x = temp_mat[, 1:6],
  argvar = list(fun = "poly", degree = 2),
  arglag = list(fun = "cr")
  )

temp_k_lin <-
  crossbasis(
    x = temp_mat[, 1:6],
    argvar = list(fun = "lin"),
    arglag = list(fun = "cr")
  )
# summary(temp_k)
# summary(temp_k_lin)


precip_k <-
  crossbasis(
  x = precip_mat[, 1:6],
  argvar = list(fun = "poly", degree = 2),
  arglag = list(fun = "cr")
  )

precip_k_lin <-
  crossbasis(
    x = precip_mat[, 1:6],
    argvar = list(fun = "lin"),
    arglag = list(fun = "cr")
  )

# summary(precip_k)
```


## default penalties
```{r}
temp_pen <- cbPen(temp_k)
precip_pen <- cbPen(precip_k)

temp_pen_lin <- cbPen(temp_k_lin)
precip_pen_lin <- cbPen(precip_k_lin)
```

# Include both temp and precip in the same model

```{r}
w_gam <-
  gam(growth ~ temp_k + precip_k + day_post + diameter,
      data = shootdf,
      paraPen = list(precip_k = precip_pen, temp_k = temp_pen),
      gamma = 1.2)
anova.gam(w_gam)
# summary(w_gam)
```

Check co-variates
```{r}
w_gam2 <-
  gam(growth ~ precip_k + temp_k + day_post,
      data = shootdf,
      paraPen = list(precip_k = precip_pen, temp_k = temp_pen),
      gamma = 1.2)

w_gam3 <-
  gam(growth ~ precip_k + temp_k + diameter,
      data = shootdf,
      paraPen = list(precip_k = precip_pen, temp_k = temp_pen),
      gamma = 1.2)

w_gam4 <-
  gam(growth ~ precip_k + temp_k,
      data = shootdf,
      paraPen = list(precip_k = precip_pen, temp_k = temp_pen),
      gamma = 1.2)

ICtab(w_gam, w_gam2, w_gam3, w_gam4)
```
Co-variates all matter

check main effects shape

```{r}
w_gam5 <-
  gam(growth ~ precip_k_lin + temp_k + day_post + diameter,
      data = shootdf,
      paraPen = list(precip_k_lin = precip_pen_lin, temp_k = temp_pen),
      gamma = 1.2)

w_gam6 <-
  gam(growth ~ precip_k + temp_k_lin + day_post + diameter,
      data = shootdf,
      paraPen = list(precip_k = precip_pen, temp_k_lin = temp_pen_lin),
      gamma = 1.2)

w_gam7 <-
  gam(growth ~ precip_k_lin + temp_k_lin + day_post + diameter,
      data = shootdf,
      paraPen = list(precip_k_lin = precip_pen_lin, temp_k_lin = temp_pen_lin),
      gamma = 1.2)

ICtab(w_gam, w_gam5, w_gam6, w_gam7)
```
Maybe precip should be linear.

Test for significance of main effects

```{r}
w_gam8 <-
  gam(growth ~temp_k + day_post + diameter,
      data = shootdf,
      paraPen = list(temp_k = temp_pen),
      gamma = 1.2)

w_gam9 <-
  gam(growth ~ precip_k_lin + day_post + diameter,
      data = shootdf,
      paraPen = list(precip_k_lin = precip_pen_lin),
      gamma = 1.2)

w_lm10 <-
  lm(growth ~ day_post + diameter,
      data = shootdf)

ICtab(w_gam5, w_gam8, w_gam9, w_lm10)
```

Winner:

```{r}
w_gam
```

Both significant

```{r}
pred.temp <- crosspred(
  temp_k, #the lagged predictor variable
  w_gam, #the glm model
  # cen = 24,
  # at=0:20, #what values of PM10 to evaluate at
  bylag=0.2, #how fine to increment the lag space.  Small number makes smoother
  cumul=TRUE #"incremental cumulative associations along lags are also predicted"
  )

pred.precip <- crosspred(
  precip_k, #the lagged predictor variable
  w_gam, #the glm model
  # at=0:20, #what values of PM10 to evaluate at
  bylag=0.2, #how fine to increment the lag space.  Small number makes smoother
  cumul=TRUE #"incremental cumulative associations along lags are also predicted"
  )
```

```{r}
plot(pred.temp, "contour",
     ylab = "lag (days)",
     xlab = "mean temperature (ºC)")
```
```{r}
plot(pred.precip, "contour",
     ylab = "lag (days)",
     xlab = "precip (mm)")
```




Look at coefs

```{r}
coef(w_gam) %>%
  enframe() %>% 
  filter(str_starts(name, "temp_kv1")) %>% 
  separate(name, c("label", "coef"), sep = "\\.l") %>% 
  mutate(lag = seq(0, 5, length.out = n())) %>% 
  ggplot(aes(x = lag, y = value)) +
  geom_point() +
  geom_line() +
  ggtitle("Temperature coefs")

summary(temp_k)
```

```{r}
coef(w_gam) %>%
  enframe() %>% 
  filter(str_starts(name, "precip_kv2")) %>% 
  separate(name, c("label", "coef"), sep = "\\.l") %>% 
  mutate(lag = seq(0, 5, length.out = n())) %>% 
  ggplot(aes(x = lag, y = value)) +
  geom_point() +
  geom_line() +
  ggtitle("Precip coefs")
```

# Conclusion

There is a significant effect of temperature on tea shoot growth.  A distributed lag model shows the same thing, and is not really conceptually necessary for this analysis.  I should probably only report the results of the LMER model.

For precipitation, a 1-day lag has no effect on growth.  The distributed lag model is highly significant though showing a quadratic effect of prcip 3 days ago.